---
title: "DAM Final Project"
author: "Emmy-Sophie Herzog og Liv Belert Mortensen"
date: "26-5-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyverse)
library(dplyr)
library(readr)
library(gridExtra)

```

## Marriage census in 1845

```{r using readr for reading .csv file, and dplyr for inspecting the document}

marriage_census <- read_csv2("data/census_1845_fullycleaned_final.csv.csv", is.na(NA))

tibble(marriage_census)

```
## Investigating how many married and unmarried women are in the age group 20-30 in the 1845 census

```{r summarizing marital status using group_by() and summarise() dplyr and tidyverse functions}

marriage_summary <- marriage_census %>%
  group_by(marital_status) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(marriage_summary)

```

```{r creating a bar plot of the marital status summary with ggplot2}

ggplot(marriage_summary, aes(x = marital_status, y = count, fill = marital_status)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Women by Marital Status (Aarhus, 1845)",
       x = "Marital Status",
       y = "Number of Women",
       fill = "Marital Status") +
  scale_y_continuous(breaks = seq(0, max(marriage_summary$count), 250)) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

```


## What is the ratio of married to unmarried women in the selected age group 20-30 in 1845?

```{r calculating the ratio of married and unmarried women}

married_count <- 954
unmarried_count <- 3094

married_to_unmarried_ratio <- married_count / unmarried_count

married_to_unmarried_ratio

round(married_to_unmarried_ratio, 2)

```

```{r creating a pie chart showing the ratio of marital status with the count from the chunk above with ggplot2}

marital_data <- data.frame(
  status = c("Married", "Unmarried"),
  count = c(954, 3094))

marital_data$percentage <- round(marital_data$count / sum(marital_data$count) * 100, 1)
marital_data$label <- paste0(marital_data$status, " (", marital_data$percentage, "%)")

pie_chart_ratio <- ggplot(marital_data, aes(x = "", y = count, fill = status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  labs(title = "Ratio of Married to Unmarried Women (Aarhus, 1845)",
       fill = "Marital Status") +
  theme_void(base_size = 14)

pie_chart_ratio

```

## Marital status in relation to occupation

What are your expectations in this regard, and how does the data correspond to your expectations?
Can you explore the reasons behind marriage age statistics / what kinds of women are getting married early vs late? (family background, status, etc.) 
we suspect that there is a relation between being unmarried vs. married. Since poorer families could not afford to lose the extra help for farms, etc.

```{r 7 most common occupations overall, calculating the top 7 occupations and creating a bar plot}

top_occupations_all <- marriage_census %>%
  count(occupation, sort = TRUE) %>%
  slice_max(n, n = 7)

plot_top_7_all <- ggplot(top_occupations_all, aes(x = reorder(occupation, n), y = n, fill = occupation)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 7 Occupations of Women ages 20-30 (Aarhus, 1845)",
       x = "Occupation",
       y = "Number of Women") +
  theme_minimal(base_size = 12)

plot_top_7_all

```

```{r top 7 occupations of unmarried women, filtering for unmarried unwomen and creating a bar plot}

top_occupations_unmarried <- marriage_census %>%
  filter(marital_status == "ugift") %>%
  count(occupation, sort = TRUE) %>%
  slice_max(n, n = 7)

plot_top_7_unmarried <- ggplot(top_occupations_unmarried, aes(x = reorder(occupation, n), y = n, fill = occupation)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 7 Occupations of Unmarried Women ages 20-30 (Aarhus, 1845)",
       x = "Occupation",
       y = "Number of Women") +
  theme_minimal(base_size = 11)

plot_top_7_unmarried

```

```{r top 7 occupations of married women, filtering for married women and creating a bar plot}

top_occupations_married <- marriage_census %>%
  filter(marital_status == "gift") %>%
  count(occupation, sort = TRUE) %>%
  slice_max(n, n = 7)

plot_top_7_married <- ggplot(top_occupations_married, aes(x = reorder(occupation, n), y = n, fill = occupation)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 7 Occupations of Married Women Ages 20-30 (Aarhus, 1845)",
       x = "Occupation",
       y = "Number of Women") +
  theme_minimal(base_size = 11)

plot_top_7_married

```

```{r using gridExtra to show the top occupations side by side for a nice visual}

grid_visuals <- grid.arrange(plot_top_7_married, plot_top_7_unmarried, ncol = 1)

```


## Average age of marriage, and age distribution for married women in their 20s in Aarhus in 1845

```{r filtering for married women and counting for each age, calculating the average and creating a bar plot with dplyt and ggplot2}

age_distribution <- marriage_census %>%
  filter(marital_status == "gift") %>%
  count(age)

average_age <- marriage_census %>%
  filter(marital_status == "gift") %>%
  summarise(avg_age = mean(age, na.rm = TRUE)) %>%
  pull(avg_age)

average_age_and_marriage_distribution <- ggplot(age_distribution, aes(x = age, y = n)) +
  geom_col(fill = "skyblue") +
  geom_vline(xintercept = average_age, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = average_age, y = max(age_distribution$n), 
           label = paste("Avg Age:", round(average_age, 1)), 
           vjust = -0.5, hjust = -0.1, color = "red", size = 4) +
  labs(title = "Marriage Age Distribution With Average Age of Marriage",
       x = "Age",
       y = "Number of Married Women Ages 20-30") +
  theme_minimal(base_size = 12)

average_age_and_marriage_distribution

```
